---
title: "Application 3: Diabetes"
format: html
execute:
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup
source("_common.R")
```

## Data Loading

```{r}
library("care")
data(efron2004, package = "care")
dim(efron2004$x) # 442 10
colnames(efron2004$x)

x_mat <- as.matrix(efron2004$x)   
class(x_mat) <- "matrix"        

X <- x_mat[,-2]         # remove "sex" (a binary variable)
y <- efron2004$y[,1]

fit_data <- list(
  N=nrow(X),
  y = y,
  P = ncol(X), 
  X = apply(X, 2, rescale_to_range1), 
  M = 40
)

```

## Exploratory Visualizations

Marginal scatter plots and GAM smooth for each predictor -\> inspect nonlinear relationships. The blue curve is a mgcv::gam fit (`y ~ s(x, bs = 'cs')`, a penalized cubic spline basis) with a 95% pointwise confidence band. Each facet fits its own smoother.

```{r}
df <- as.data.frame(cbind(y, X))
df |> 
  gather(key, val, -y) %>%
  ggplot(aes(x = val, y = y)) +
  geom_point() +
  stat_smooth(method = "gam", se = TRUE, col = "blue") +
  facet_wrap(~key, scales = "free") +
  theme_gray() +
  ggtitle("Marginal scatter plot") 
```

## Model Fitting
### independent priors on standardized coefficients theta
```{r}
fit_a3_indep <- fit_model(
        gp_add_p1_13,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
)
# saveRDS(fit_a3_indep, file = "analysis/application/fits/fit_a3_indep.rds")
# fit_a3_indep <- readRDS("analysis/application/fits/fit_a3_indep.rds")
```

### dependent prior on theta
```{r}
fit_a3_dep <- fit_model(
        gp_add_p1_15,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
)
# saveRDS(fit_a3_dep, file = "analysis/application/fits/fit_a3_dep.rds")
# fit_a3_dep <- readRDS("analysis/application/fits/fit_a3_dep.rds")
```

### non-local priors on delta

```{r}
fit_a3_dep_p2 <- fit_model(
        gp_add_p2_15,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
)
# saveRDS(fit_a3_dep_p2, file = "analysis/application/fits/fit_a3_dep_p2.rds")
# fit_a3_dep_p2 <- readRDS("analysis/application/fits/fit_a3_dep_p2.rds")
```

## Results & Diagnostics
```{r}
print_quick_summary(fit_a3_indep)
print_quick_summary(fit_a3_dep, key_params = c("delta", "lambda", "sigma", "mu", "beta", "theta"))
print_quick_summary(fit_a3_dep_p2, key_params = c("delta2", "lambda", "sigma", "mu", "beta1", "theta"))

summary_gp_fit(fit_a3_dep)
summary_gp_fit(fit_a3_dep_p2, X_orig = X, delta_prior = "p2")
```


```{r}
plot_posterior_delta(fit_a3_dep, bins=50, plot_prior=TRUE, ylim=c(0,0.8))
plot_posterior_theta(fit_a3_dep, plot_prior=TRUE, bins=50, ylim=c(0,1.6))


plot_gp_trends(fit_a3_indep, model_name="exact-GP / marginal-θ")
plot_gp_trends(fit_a3_dep, X_orig=X, ylim=c(-2,3), 
                model_name="exact-GP / multivariate-θ",)
plot_gp_trends(fit_a3_dep_p2, X_orig=X, ylim=c(-2,3), delta_prior = "p2",
                model_name="multivariate-θ / 2nd-order non-local prior")
```  

## BF: Savage-Dickey density ratios
```{r}
compute_savage_dickey_ratios(fit_a3_indep)
compute_savage_dickey_ratios(fit_a3_dep)
compute_savage_dickey_ratios(fit_a3_dep_p2, delta_prior = "p2")
```


=========================================================================================
## HSGP approximation
```{r}
time_hsgp_a3 <- system.time({
fit_a3_indep_hsgp <- fit_model(
        gp_hsgp_02,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
      )
})

fit_a3_dep_hsgp <- fit_model(
        gp_hsgp_01,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
      )

```

```{r}
print_quick_summary(fit_a3_indep_hsgp)
print_quick_summary(fit_a3_dep_hsgp)
```

```{r}
compute_savage_dickey_ratios(fit_a3_indep_hsgp)
compute_savage_dickey_ratios(fit_a3_dep_hsgp)
```

```{r}
plot_hsgp_trends(fit_a3_indep_hsgp, model_name="HSGP / marginal-θ")
plot_hsgp_trends(fit_a3_dep_hsgp, model_name="HSGP / multivariate-θ", ylim=c(-3,3))
```

=========================================================================================
## Compare with GAM
```{r}
library(mgcv)

# Fit GAM with same structure as GP model: y = linear + smooth terms
# This matches the GP structure: y = beta0 + X*beta1 + f1(x1) + ... + f9(x9)

# Create data frame for GAM
gam_data <- data.frame(
  y = y,
  X  # X is already scaled from fit_data
)
# Make sure column names match the original predictor names
colnames(gam_data)[-1] <- colnames(efron2004$x)[-2]  # Remove "sex" column

# Fit GAM with linear + smooth terms for each predictor
# s(xi, fx=FALSE) allows the smoother to determine its own complexity
gam_fit_time <- system.time({
  gam_fit <- gam(y ~ age + bmi + bp + s1 + s2 + s3 + s4 + s5 + s6 +
                     s(age) + s(bmi) + s(bp) + s(s1) + s(s2) + 
                     s(s3) + s(s4) + s(s5) + s(s6),
                 data = gam_data,
                 method = "REML")
})

# Summary of GAM fit
summary(gam_fit)

# Display time taken
gam_fit_time
```

- plot see `/Users/ywu/Documents/phd-project-01/experiments/260107_gam_diabetes.R`






```{r}
# X: N x P matrix used in Stan
# beta1_draws: posterior draws of beta1 (S x P), e.g. from posterior::as_draws_df()

library(posterior)

draws <- as_draws_df(fit_a3_dep$fit)

# --- 1) single-number check: posterior mean of X %*% beta1 ---
P <- ncol(X)

beta1_mat <- as.matrix(draws[, paste0("beta1[", 1:P, "]")])  # S x P
beta1_mean <- colMeans(beta1_mat)                             # length P

Xbeta1_mean <- as.vector(X %*% beta1_mean)                    # length N
summary(Xbeta1_mean)

# --- 2) pointwise posterior distribution of X %*% beta1 (at each observation) ---
Xbeta1_draws <- beta1_mat %*% t(X)    # S x N (since (SxP)*(PxN))
# posterior mean for each i
Xbeta1_point_mean <- colMeans(Xbeta1_draws)                   # length N
# posterior quantiles for each i (2.5%, 50%, 97.5%)
Xbeta1_point_q <- apply(Xbeta1_draws, 2, quantile, probs = c(.025, .5, .975))

# quick look at a few observations:
head(Xbeta1_point_mean)
Xbeta1_point_q[, 1:5]

# --- 3) scalar summary: the dot product x_bar' beta1 (this is what drives beta0) ---
x_bar <- colMeans(X)
xbar_beta1_draws <- as.vector(beta1_mat %*% x_bar)            # length S
mean(xbar_beta1_draws)
quantile(xbar_beta1_draws, c(.025, .5, .975))

# If you used centered X in Stan, and you want X_c %*% beta1 instead:
X_c <- scale(X, center = TRUE, scale = FALSE)
Xcbeta1_mean <- as.vector(X_c %*% beta1_mean)
summary(Xcbeta1_mean)

```


```{r}
# posterior mean of beta1
beta1_mean <- colMeans(beta1_mat)

x_bar <- colMeans(X)

# this should be approximately -beta0_mean (since y_bar ~ 0)
xbar_beta1 <- sum(x_bar * beta1_mean)
xbar_beta1

mean_beta0 <- mean(draws$beta0)
mean_beta0

mean_beta0 + xbar_beta1  # should be ~0 (since y_bar ~ 0)

```



```{r}
library(posterior)

draws <- as_draws_df(fit_a3_dep$fit)

P <- ncol(X)

# Extract beta1 draws: S x P
beta1_mat <- as.matrix(draws[, paste0("beta1[", 1:P, "]")])

# Recompute x_bar from the EXACT X you passed into Stan
x_bar <- colMeans(X)

# Stan's dot_product(x_bar, beta1) per draw: length S
xbar_beta1_draw <- as.vector(beta1_mat %*% x_bar)

# Now check the identity draw-by-draw.
# y_bar in Stan is mean(y). You computed mean(y) ~ 0, use that:
y_bar <- mean(y)

beta0_draw <- draws$beta0

xbar_beta1_draw2 <- rowSums(beta1_mat * rep(x_bar, each=nrow(beta1_mat)))  # S
summary(y_bar - xbar_beta1_draw2)


summary(beta0_draw)
summary(-xbar_beta1_draw)          # should match beta0_draw if y_bar ~ 0
summary(y_bar - xbar_beta1_draw)
summary(beta0_draw - (y_bar - xbar_beta1_draw))

max(abs(beta0_draw - (y_bar - xbar_beta1_draw)))

```


```{r}
draws <- posterior::as_draws_df(fit_a3_dep$fit)
y_bar <- mean(y)  # ~0
xbar_beta1_implied <- y_bar - draws$beta0
summary(xbar_beta1_implied)
```


```{r}
d <- fit_a3_dep$data
X_stan <- d$X
y_stan <- d$y

c(mean(y_stan), mean(fit_data$y))             # should match
range(colMeans(X_stan))              # THIS determines x_bar in Stan
range(colMeans(fit_data$X))                   # compare to your current X
```






```{r}
library(posterior)

fit <- fit_a3_dep$fit
draws <- as_draws_df(fit)

# Extract exactly the data used by Stan
d <- fit_a3_dep$data
X_used <- d$X
y_used <- d$y

P <- ncol(X_used)

beta1_names <- paste0("beta1[", 1:P, "]")
beta1_mat <- as.matrix(draws[, beta1_names])

x_bar_used <- colMeans(X_used)
y_bar_used <- mean(y_used)

# dot_product(x_bar, beta1) per draw
xbar_beta1_draw <- as.vector(beta1_mat %*% x_bar_used)

# This MUST match beta0 (draw-by-draw) if beta0 was defined as y_bar - dot_product(...)
beta0_recon <- y_bar_used - xbar_beta1_draw

summary(draws$beta0)
summary(beta0_recon)

summary(draws$beta0 - beta0_recon)
max(abs(draws$beta0 - beta0_recon))

```



```{r}

for ( i in 1:1e3 ) {
  beta1 <- beta1_mat[i, ]
  xbar_beta1 <- sum(x_bar_used * beta1)
  beta0_recon <- y_bar_used - xbar_beta1
}
```



```{r}
X_data <- fit_data$X - colMeans(fit_data$X)
coef_lm <- as.vector(lm(y~X_data)$coef)
beta1 <- coef_lm[2:10]
mean(y) - sum(beta1 * mean(X_data))
```

original X: -1.175096e-16
centered original X: -1.651289e-16
rescaled X: 0.4482359
centered rescaled X: 4.531069e-17
