---
title: "Application 3: Diabetes"
format: html
execute:
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup
source("_common.R")
```

## Data Loading

```{r}
library("care")
data(efron2004)
dim(efron2004$x) # 442 10
colnames(efron2004$x)

x_mat <- as.matrix(efron2004$x)   
class(x_mat) <- "matrix"        

X <- x_mat[,-2]         # remove "sex" (a binary variable)
y <- efron2004$y[,1]

fit_data <- list(
  N=nrow(X),
  y = y,
  P = ncol(X), 
  X = apply(X, 2, rescale_to_range1), 
  M = 40
)
```

## Exploratory Visualizations

Marginal scatter plots and GAM smooth for each predictor -\> inspect nonlinear relationships. The blue curve is a mgcv::gam fit (`y ~ s(x, bs = 'cs')`, a penalized cubic spline basis) with a 95% pointwise confidence band. Each facet fits its own smoother.

```{r}
df <- as.data.frame(cbind(y, X))
df |> 
  gather(key, val, -y) %>%
  ggplot(aes(x = val, y = y)) +
  geom_point() +
  stat_smooth(method = "gam", se = TRUE, col = "blue") +
  facet_wrap(~key, scales = "free") +
  theme_gray() +
  ggtitle("Marginal scatter plot") 
```

## Model Fitting
### independent priors on standardized coefficients theta
```{r}
fit_a3_indep <- fit_model(
        gp_add_p1_13,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
)
# saveRDS(fit_a3_indep, file = "analysis/application/fits/fit_a3_indep.rds")
# fit_a3_indep <- readRDS("analysis/application/fits/fit_a3_indep.rds")
```

### dependent prior on theta
```{r}
fit_a3_dep <- fit_model(
        gp_add_p1_14,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
)
# saveRDS(fit_a3_dep, file = "analysis/application/fits/fit_a3_dep.rds")
# fit_a3_dep <- readRDS("analysis/application/fits/fit_a3_dep.rds")
```

## Results & Diagnostics
```{r}
print_quick_summary(fit_a3_indep)
print_quick_summary(fit_a3_dep, key_params = c("delta", "lambda", "sigma", "mu", "beta"))
```


```{r}
plot_posterior(fit_a3_indep)
plot_posterior(fit_a3_dep)

plot_gp_trends(fit_a3_indep, model_name="exact-GP / marginal-θ")
plot_gp_trends(fit_a3_dep, model_name="exact-GP / multivariate-θ",ylim=c(-3,3))
```  

## BF: Savage-Dickey density ratios
```{r}
compute_savage_dickey_ratios(fit_a3_indep)
compute_savage_dickey_ratios(fit_a3_dep)
```


=========================================================================================
## HSGP approximation
```{r}
time_hsgp_a3 <- system.time({
fit_a3_indep_hsgp <- fit_model(
        gp_hsgp_02,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
      )
})

fit_a3_dep_hsgp <- fit_model(
        gp_hsgp_01,
        data = fit_data,
        chains = 2,
        iter = 5500,
        warmup = 500
      )

```

```{r}
print_quick_summary(fit_a3_indep_hsgp)
print_quick_summary(fit_a3_dep_hsgp)
```

```{r}
compute_savage_dickey_ratios(fit_a3_indep_hsgp)
compute_savage_dickey_ratios(fit_a3_dep_hsgp)
```

```{r}
plot_hsgp_trends(fit_a3_indep_hsgp, model_name="HSGP / marginal-θ")
plot_hsgp_trends(fit_a3_dep_hsgp, model_name="HSGP / multivariate-θ", ylim=c(-3,3))
```

=========================================================================================
## Compare with GAM
```{r}
library(mgcv)

# Fit GAM with same structure as GP model: y = linear + smooth terms
# This matches the GP structure: y = beta0 + X*beta1 + f1(x1) + ... + f9(x9)

# Create data frame for GAM
gam_data <- data.frame(
  y = y,
  X  # X is already scaled from fit_data
)
# Make sure column names match the original predictor names
colnames(gam_data)[-1] <- colnames(efron2004$x)[-2]  # Remove "sex" column

# Fit GAM with linear + smooth terms for each predictor
# s(xi, fx=FALSE) allows the smoother to determine its own complexity
gam_fit_time <- system.time({
  gam_fit <- gam(y ~ age + bmi + bp + s1 + s2 + s3 + s4 + s5 + s6 +
                     s(age) + s(bmi) + s(bp) + s(s1) + s(s2) + 
                     s(s3) + s(s4) + s(s5) + s(s6),
                 data = gam_data,
                 method = "REML")
})

# Summary of GAM fit
summary(gam_fit)

# Display time taken
gam_fit_time
```

- plot see `/Users/ywu/Documents/phd-project-01/experiments/260107_gam_diabetes.R`


```{r}
bridgesampling::bridge_sampler(try_fit)
bridgesampling::bridge_sampler(fit_a3_indep_hsgp$fit)
```